GLOBALIZER: A NOVEL SUPERCOMPUTER SOFTWARE SYSTEM FOR SOLVING TIME-CONSUMING GLOBAL OPTIMIZATION PROBLEMS

### 2016-08-10 2016/GOM/paper.tex
Update references in GOM paper

optimization methods --- see, for example, (Gergel, V. P., 1996, 1997), (Gergel, V. P., Strongin, R. G., 2003, 2005), (Grishagin, V. A., Strongin, R. G., 1984), (Sergeyev Y. D., 1995, 1999), (Sergeyev, Y. D., Grishagin V. A., 1994, 2001), (Sergeyev, Y. D., Strongin, R. G., Lera, D., 2013), (Barkalov K. A., Gergel V. P., 2014), etc.

@@

instead of applying a single Peano curve \(y(x)\) has been proposed in (Strongin, R. G., 1992, \cite{strSergGO}).

@@

Thus, for example, in (Strongin, R. G., 1992) a scheme has been applied by which each

@@

Additional information on organizing such parallel computation schemes is provided in (Gergel, V. P., Sidorov, S., 2015).

\section{Globalizer System Architecture}
\label{sec:arch}
As previously mentioned in Section \ref{sec:intro}, the development of software systems
for global optimization is complex professional work. The optimization systems should
include advanced optimization methods, provide for the accumulation and efficient
processing of search information, include various forms of visualization for the results,
provide tools for interacting with users, etc. All the features pointed out above
become much more complicated when organizing parallel computations using high-performance
multiprocessor systems of varying architectures.
\par
The Globalizer considered in this paper expands the family of global optimization
software systems successively developed by the authors during the past several years.
One of the first developments was the SYMOP multiextremal optimization system (see Gergel, V. P., 1993),
which has been successfully applied for solving many optimization problems. A special
place is occupied by the ExaMin system (see, for example, (Barkalov, K., Gergel V., 2015)),
which was developed and used extensively to investigate the application of novel parallel
algorithms to solve global optimization problems using high-performance multiprocessor computing systems.

@@

The Globalizer system architecture is presented in Figure \ref{fig:globalizerScheme}. The structural components are:
  \item Block 0 consists of the procedures for computing the function values (criteria
  and constraints) for the optimization problem being solved; this is an external
  block with respect to the Globalizer system and is provided by the user per a predefined interface.
  \item Blocks 1-4 form the optimization subsystem and solve the global optimization
  problems (Block 1), nonlinear programming (Block 2), multicriterial optimization (Block 3),
  and general decision making problems (Block 4). It is worth noting the sequential
  scheme of interaction between these components --- the decision making problems are
  solved using the multicriterial optimization block, which, in turn, uses the nonlinear programming block, etc.
  \item Block 5 is a subsystem for accumulating and processing the search information;
  this is one of the basic subsystems --- the amount of search information for complex
  optimization problems may appear to be quite large on the one hand, but, on the other
  hand, the efficiency of the global optimization methods depends to a great extent on how
  completely all of the available search data is utilized.
  \item Block 6 contains the dimensional reduction procedures based on Peano evolvents;
  the optimization blocks solve the reduced (one-dimensional) optimization problems;
  Block 6 provides interaction between the optimization blocks and the initial
  multidimensional optimization problem.
  \item Block 7 organizes the choice of parallel computation schemes in the Globalizer
  system subject to the computing system architecture employed (the numbers of cores in the
  processors, the availability of shared and distributed memory, the availability of
  accelerators for computations, etc.) and the global optimization methods applied.
  \item Block 8 is responsible for managing the parallel processes when performing the
  global search (determining the optimal configuration of parallel processes, distributing
  the processes between computer elements, balancing computational loads, etc.).
  \item Block 9 is a common managing subsystem, which fully controls the whole
  computational process when solving global optimization problems.
  \item Block 10 is responsible for organizing the dialog interaction with users for
  stating the optimization problem, adjusting system parameters (if necessary), and
  visualizing and presenting the global search results.
  \item Block 11 is a set of tools for visualizing and presenting the global search results;
  the availability of tools for visually presenting the computational results enables
  the user to provide efficient control over the global optimization process.
The Globalizer system comes in two different options:
  \item Globalizer Pro is a research version of the system providing a full set of
  capabilities that are oriented toward solving of the most complicated global optimization problems.
  \item Globalizer Lite is a limited version of the Globalizer system accessible for free for
  solving global optimization problems with a medium level of difficulty, and is accessible free of charge.
\section{Results of Numerical Experiments}
\label{sec:experiments}
In order to confirm the operability of the Globalizer system, results from two
series of computational experiments are presented in this section.
The computational experiments were conducted using the Lobachevsky supercomputer at the
State University of Nizhny Novgorod (operation system --- CentOS 6.4, management system --- SLURM).
Each supercomputer node had 2 Intel Sandy Bridge E5-2660 processors, 2.2 GHz, with 64 GB RAM.
Each processor had 8 cores (i.e. a total of 16 CPU cores were available at each node).
In order to generate the executable program code, the Intel C++ 14.0.2 compiler was used.
A number of nodes in the computer system utilized Intel Xeon Phi 5110P processors,
each containing 60 cores (a total of 240 threads). In addition, the supercomputer also
included computer nodes equipped with two GPU NVIDIA Kepler K20X, each providing 14
stream multiprocessors (a total of 2688 CUDA-cores). In order to generate the executable
code, the CUDA Toolkit 6.0 was used.
The problems generated by the GKLS-generator (Gaviano, M., Lera, D., Kvasov, D. E., Sergeyev, Y. D., 2003)
were selected for the test problems. This generator allowed for obtaining the multiextremal
optimization problems with \textit{a priori} known properties: the number of local minima,
the sizes of their attractors, the global minimum point, the function value at this point, etc.
In order to simulate the computation costs inherent in the applied optimization problems
(see, for example, Bastrakov, S., Gergel, V., Meyerov, I., et al, 2013), in all of the experiments
computing the objective function was complicated by auxiliary computations, not altering the
form of the function and the position of its minima.
We considered the global minimum \(y^*\) in the test problem to be found if the
algorithm generated a trial point \(y^k\) in the \(\delta\)-nearness of the global
minimum i. e. \(\Vert y^k-y^* \Vert \leqslant \delta\). The size of the nearness was
selected as \(\delta=\Vert b-a \Vert\sqrt[N]{\Delta}\) where \(a\) and \(b\) are
boundaries of the search domain \(D\), \(N\) is the dimensionality of the optimization problem.
The maximum allowed number of parallel iterations was \(K_{max} = 10000\), and the evolvent
density parameter was \(m=10\).The computational experiments considered below were first
carried out using the ExaMin system (see (Barkalov, K., Gergel V., 2015), (Barkalov, K., Gergel, V., Lebedev, I., 2015), (Barkalov, K., Gergel, V., Lebedev, I., Sysoev, A., 2015), (Gergel, V., Lebedev, I., 2015))
and then were reproduced using the Globalizer system.

>>>>

optimization methods --- see, for example, \cite{gergel1996}, \cite{gergel1997},
\cite{gergelStrongin2003}, \cite{gergelStrongin2005}, \cite{grishaginStrongin1984},
\cite{sergeyev1995}, \cite{sergeyev1999}, \cite{sergeyevGrishagin1994}, \cite{sergeyevGrishagin2001}
\cite{sergeyevStronginLera2013}, \cite{barkalovGergel2014}, etc.

@@

instead of applying a single Peano curve \(y(x)\) has been proposed in \cite{strongin1992}, \cite{strSergGO}.

@@

Thus, for example, in \cite{strongin1992} a scheme has been applied by which each

@@

Additional information on organizing such parallel computation schemes is provided in \cite{gergelSidorov2015}.

\section{Globalizer System Architecture}
\label{sec:arch}
As previously mentioned in Section \ref{sec:intro}, the development of software systems
for global optimization is complex professional work. The optimization systems should
include advanced optimization methods, provide for the accumulation and efficient
processing of search information, include various forms of visualization for the results,
provide tools for interacting with users, etc. All the features pointed out above
become much more complicated when organizing parallel computations using high-performance
multiprocessor systems of varying architectures.
\par
The Globalizer considered in this paper expands the family of global optimization
software systems successively developed by the authors during the past several years.
One of the first developments was the SYMOP multiextremal optimization system \cite{gergel1993},
which has been successfully applied for solving many optimization problems. A special
place is occupied by the ExaMin system (see, for example, \cite{barkalovGergel2015}),
which was developed and used extensively to investigate the application of novel parallel
algorithms to solve global optimization problems using high-performance multiprocessor computing systems.

@@

\par
The Globalizer system architecture is presented in Figure \ref{fig:globalizerScheme}. The structural components are:
\begin{itemize}
  \item Block 0 consists of the procedures for computing the function values (criteria
  and constraints) for the optimization problem being solved; this is an external
  block with respect to the Globalizer system and is provided by the user per a predefined interface.
  \item Blocks 1-4 form the optimization subsystem and solve the global optimization
  problems (Block 1), nonlinear programming (Block 2), multicriterial optimization (Block 3),
  and general decision making problems (Block 4). It is worth noting the sequential
  scheme of interaction between these components --- the decision making problems are
  solved using the multicriterial optimization block, which, in turn, uses the nonlinear programming block, etc.
  \item Block 5 is a subsystem for accumulating and processing the search information;
  this is one of the basic subsystems --- the amount of search information for complex
  optimization problems may appear to be quite large on the one hand, but, on the other
  hand, the efficiency of the global optimization methods depends to a great extent on how
  completely all of the available search data is utilized.
  \item Block 6 contains the dimensional reduction procedures based on Peano evolvents;
  the optimization blocks solve the reduced (one-dimensional) optimization problems;
  Block 6 provides interaction between the optimization blocks and the initial
  multidimensional optimization problem.
  \item Block 7 organizes the choice of parallel computation schemes in the Globalizer
  system subject to the computing system architecture employed (the numbers of cores in the
  processors, the availability of shared and distributed memory, the availability of
  accelerators for computations, etc.) and the global optimization methods applied.
  \item Block 8 is responsible for managing the parallel processes when performing the
  global search (determining the optimal configuration of parallel processes, distributing
  the processes between computer elements, balancing computational loads, etc.).
  \item Block 9 is a common managing subsystem, which fully controls the whole
  computational process when solving global optimization problems.
  \item Block 10 is responsible for organizing the dialog interaction with users for
  stating the optimization problem, adjusting system parameters (if necessary), and
  visualizing and presenting the global search results.
  \item Block 11 is a set of tools for visualizing and presenting the global search results;
  the availability of tools for visually presenting the computational results enables
  the user to provide efficient control over the global optimization process.
\end{itemize}
\par
The Globalizer system comes in two different options:
\begin{itemize}
  \item Globalizer Pro is a research version of the system providing a full set of
  capabilities that are oriented toward solving of the most complicated global optimization problems.
  \item Globalizer Lite is a limited version of the Globalizer system accessible for free for
  solving global optimization problems with a medium level of difficulty, and is accessible free of charge.
\end{itemize}

\section{Results of Numerical Experiments}
\label{sec:experiments}
In order to confirm the operability of the Globalizer system, results from two
series of computational experiments are presented in this section.
\par
The computational experiments were conducted using the Lobachevsky supercomputer at the
State University of Nizhny Novgorod (operation system --- CentOS 6.4, management system --- SLURM).
Each supercomputer node had 2 Intel Sandy Bridge E5-2660 processors, 2.2 GHz, with 64 GB RAM.
Each processor had 8 cores (i.e. a total of 16 CPU cores were available at each node).
In order to generate the executable program code, the Intel C++ 14.0.2 compiler was used.
\par
A number of nodes in the computer system utilized Intel Xeon Phi 5110P processors,
each containing 60 cores (a total of 240 threads). In addition, the supercomputer also
included computer nodes equipped with two GPU NVIDIA Kepler K20X, each providing 14
stream multiprocessors (a total of 2688 CUDA-cores). In order to generate the executable
code, the CUDA Toolkit 6.0 was used.
\par
The problems generated by the GKLS-generator \cite{gavianoKvasovLeraSergeev2003}
were selected for the test problems. This generator allowed for obtaining the multiextremal
optimization problems with \textit{a priori} known properties: the number of local minima,
the sizes of their attractors, the global minimum point, the function value at this point, etc.
In order to simulate the computation costs inherent in the applied optimization problems
(see, for example, Bastrakov, S., Gergel, V., Meyerov, I., et al, 2013), in all of the experiments
computing the objective function was complicated by auxiliary computations, not altering the
form of the function and the position of its minima.
\par
We considered the global minimum \(y^*\) in the test problem to be found if the
algorithm generated a trial point \(y^k\) in the \(\delta\)-nearness of the global
minimum i. e. \(\Vert y^k-y^* \Vert \leqslant \delta\). The size of the nearness was
selected as \(\delta=\Vert b-a \Vert\sqrt[N]{\Delta}\) where \(a\) and \(b\) are
boundaries of the search domain \(D\), \(N\) is the dimensionality of the optimization problem.
The maximum allowed number of parallel iterations was \(K_{max} = 10000\), and the evolvent
density parameter was \(m=10\). The computational experiments considered below were first
carried out using the ExaMin system (see \cite{barkalovGergel2015}, \cite{barkalovGergelLebedev2015},
\cite{barkalovGergelLebedevSysoev2015}, \cite{gergelLebedev2015})
and then were reproduced using the Globalizer system.
